---
title: "create CD3+CD8-"
author: "Ming Tang"
date: '2023-07-26'
output: html_document
---


```{r}
library(here)
library(tidyverse)
library(Seurat)
library(scattermore)
```


Read in the Seurat object

```{r}
cd3_plus_annotated<- readRDS(here("data/cd3_plus_annotated_seurat_subcluster.rds"))
```

### keep cells with 0 count of CD8A and CD8B

exclude B cells (contamination) and red blood cells

```{r}
cd3_plus_annotated@meta.data$annotation %>% table()

rownames(cd3_plus_annotated) %>% str_subset("CD8")
(cd3_plus_annotated[["RNA"]]@counts["CD8A", ] == 0) %>% table()

(cd3_plus_annotated[["RNA"]]@counts["CD8B", ] == 0) %>% table()

(cd3_plus_annotated[["RNA"]]@counts["CD8A", ] == 0 & cd3_plus_annotated[["RNA"]]@counts["CD8B", ] == 0) %>% table()

cd8_minus_index<- cd3_plus_annotated[["RNA"]]@counts["CD8A", ] == 0 & cd3_plus_annotated[["RNA"]]@counts["CD8B", ] == 0

(cd3_plus_annotated@meta.data$annotation != "Red_Blood_Cells"  & cd3_plus_annotated@meta.data$annotation != "B_cells") %>% table()

cd8_minus_index2<- cd3_plus_annotated@meta.data$annotation != "Red_Blood_Cells"  & cd3_plus_annotated@meta.data$annotation != "B_cells"

(cd8_minus_index & cd8_minus_index2) %>% table()

cd8_minus_index3<- cd8_minus_index & cd8_minus_index2

cd3_plus_cd8_minus<- cd3_plus_annotated[, cd8_minus_index3]

saveRDS(cd3_plus_cd8_minus, here("results/cd3_plus_cd8_minus.rds"))

cd3_plus_cd8_minus<- readRDS(here("results/cd3_plus_cd8_minus.rds"))
```


### submit a sbatch job.

```{bash}
#!/bin/bash
#SBATCH --job-name=cd8_minus_cluster
#SBATCH --mem=64G        # total memory need
#SBATCH -n 4 # 1 core
#SBATCH -N 1 # on one node
#SBATCH --time=12:00:00 # DD-HH:MM:SS max walltime(time limit) requested - can use -t 5:00:00 for short
#SBATCH --error=/liulab/mtang/lsf_logs/lsf_%j_%x.err      # error file
#SBATCH --output=//liulab/mtang/lsf_logs/lsf_%j_%x.err      # output file
#SBATCH --mail-type=END,FAIL # email notification when job ends/fails
#SBATCH --mail-user=mtang@ds.dfci.harvard.edu # email to notify

module load singularity

singularity exec /liulab/mtang/singularity_images/rstudio.simg Rscript cluster_recompute_PCA.R cd3_plus_cd8_minus.rds cd3_plus_cd8_minus_res1.5_recompute_Harmony_HVG.rds

```

### visualize 

```{r}

cd8_minus<- readRDS(here("results/cd3_plus_cd8_minus_res1.5.rds"))

# Free X-Y relate to Supp Figure 1
cd8_minus_recompute<- readRDS(here("results/cd3_plus_cd8_minus_res1.5_recompute_Harmony_HVG.rds"))

table(cd8_minus@meta.data$seurat_clusters)
table(cd8_minus_recompute@meta.data$seurat_clusters)

library(scclusteval)

pdf(here("results/figures/cd8_jaccard.pdf"), width = 8, height = 6)
PairWiseJaccardSetsHeatmap(Idents(cd8_minus), Idents(cd8_minus_recompute))
dev.off()

DimPlot(cd8_minus, reduction = "umap", group.by = "seurat_clusters",
        pt.size = .1, label = TRUE, repel = TRUE, order=TRUE, combine = TRUE) +
  NoLegend()

table(cd8_minus@meta.data$seurat_clusters)
table(cd3_plus_cd8_minus@meta.data$seurat_clusters)

umap_scattermore<- function(obj, reduction= "umap", col, pointsize=0) {
  scattermore_df<- cbind(obj@meta.data, Embeddings(obj, reduction = reduction))
  p<- ggplot(scattermore_df, aes(x= UMAP_1, y= UMAP_2)) + 
  geom_scattermore(aes(col= .data[[col]]), pointsize = pointsize) +
  theme_classic(base_size = 14)
  return(p)
}

umap_scattermore(cd8_minus, reduction = "umap", col = "seurat_clusters", pointsize = 0.5) +
  guides(color = guide_legend(override.aes = list(size= 5))) +
  facet_wrap(~ group, ncol = 5)
ggsave(here("results/figures/cd8_minus_subcluster_by_group.pdf"), width = 14, height = 6)


umap_scattermore(cd8_minus_recompute, reduction = "umap", col = "seurat_clusters", pointsize = 0.5) +
  guides(color = guide_legend(override.aes = list(size= 5))) +
  facet_wrap(~ group, ncol = 5)
ggsave(here("results/figures/cd8_minus_subcluster_by_group_recompute.pdf"), width = 14, height = 6)
```


```{r}
cd8_minus_recompute<- readRDS(here("results/cd3_plus_cd8_minus_res1.5_recompute_Harmony_HVG.rds"))

#cd8_minus_annotation<- read_tsv(here("data/cd8_minus_annotation.txt"))

#new cluster annotation from Julia
cd8_minus_annotation<- read_tsv(here("data/cd8_minus__annotations_20210728.txt"))
cd8_minus_annotation$seurat_clusters<- factor(cd8_minus_annotation$seurat_clusters, levels = as.character(0:32))
```


add Julia's annotation

```{r}
old_meta<- cd8_minus_recompute@meta.data
## remove the old annotation.
old_meta<- dplyr::select(old_meta, - annotation)
new_meta<- left_join(old_meta, cd8_minus_annotation)

new_meta<- as.data.frame(new_meta)
rownames(new_meta)<- rownames(old_meta)

cd8_minus_recompute@meta.data<- new_meta
```


### remove the cells that we filter out 


```{r}
indx<- !is.na(cd8_minus_recompute@meta.data$annotation)
cells_to_keep<- colnames(cd8_minus_recompute)[indx]
cd8_minus_annotated<- subset(cd8_minus_recompute, cells = cells_to_keep)
```

### subcluster 7 to 7a and 7b

There is a problem with the Matrix package https://github.com/satijalab/seurat/issues/4436

Relate to Supp Figure 3

```{r}
cd8_minus_annotated_cluster7<- subset(cd8_minus_annotated, idents = "7")

cd8_minus_annotated_cluster7<- cd8_minus_annotated_cluster7 %>%
  FindNeighbors(reduction = "harmony", dims = 1:15) %>%
  FindClusters(resolution = 0.2)

DimPlot(cd8_minus_annotated_cluster7, reduction = "umap", group.by = "seurat_clusters",
        pt.size = .1, label = TRUE, repel = TRUE, order=TRUE, combine = TRUE) +
  NoLegend()

ggsave(here("results/figures/cd8minus_subcluster7_umap.pdf"), width = 8, height = 6)

#DimPlot(cd8_minus_annotated_cluster7, reduction = "umap", group.by = "seurat_clusters",
#        pt.size = .1, label = TRUE, repel = TRUE, order=TRUE, combine = TRUE, split.by = "seurat_clusters") +
#  NoLegend()

cluster7_idents<- Idents(cd8_minus_annotated_cluster7) 
  
cluster7_a<- names(cluster7_idents)[cluster7_idents == 0]
cluster7_b<- names(cluster7_idents)[cluster7_idents == 1]

cd8_minus_annotated$clusters_anno<- paste0(str_pad(cd8_minus_annotated$seurat_clusters, width = 2, side = "left", pad = "0"), "_", cd8_minus_annotated$annotation)


anno_df<- cbind(cd8_minus_annotated@meta.data, barcode = colnames(cd8_minus_annotated)) %>%
  mutate(seurat_clusters = as.character(seurat_clusters)) %>%
  mutate(seurat_clusters_sub = case_when(
    barcode %in% cluster7_a ~ "07_a",
    barcode %in% cluster7_b ~ "07_b",
    TRUE ~ seurat_clusters
  )) %>%
  mutate(annotation_sub = case_when(
    barcode %in% cluster7_a ~ "MAIT",
    barcode %in% cluster7_b ~ "TCRgdVD2",
    TRUE ~ annotation
  )) %>%
  mutate(clusters_anno_sub =  paste0(str_pad(seurat_clusters_sub, width = 2, side = "left", pad = "0"), "_", annotation_sub))


table(anno_df$clusters_anno_sub)

cd8_minus_annotated$clusters_anno_sub<- anno_df$clusters_anno_sub
saveRDS(cd8_minus_annotated, here("data/cd8_minus_annotated_subcluster7_seurat.rds"))


cluster7_markers<- presto::wilcoxauc(cd8_minus_annotated_cluster7, 'seurat_clusters', assay = 'data')

cluster7_top_markers<- top_markers(cluster7_markers, n = 20, auc_min = 0.5, pct_in_min = 20)

cluster7_top_markers_unique<- cluster7_top_markers %>%
  select(-rank) %>% 
  unclass() %>% 
  stack() %>%
  pull(values) %>%
  unique() %>%
   .[!is.na(.)]
```


```{r}
cluster7_mat<- cd8_minus_annotated_cluster7[["RNA"]]@data[cluster7_top_markers_unique, ] %>% as.matrix()


## scale the rows
cluster7_mat<- t(scale(t(cluster7_mat)))

cluster7_cluster_anno<- cd8_minus_annotated_cluster7@meta.data$seurat_clusters


quantile(cluster7_mat, c(0.1, 0.5, 0.95))

PurpleAndYellow()
#col_fun = colorRamp2(c(-1, 0, 2), c("blue", "white", "red"))
#col_fun = colorRamp2(c(-1, 0, 2), c("#FF00FF", "black", "#FFFF00"))

col_fun<-  circlize::colorRamp2(c(-1, 0, 1.5), c("#440154FF", "#238A8DFF", "#FDE725FF"))
## for big matrix, raster helps. https://jokergoo.github.io/2020/06/30/rasterization-in-complexheatmap/
## cluster the slice (without specifiying the factor level) and the columns.
pdf(here("results/figures/cd8_minus_clusters7_marker_gene_heatmap_cluster.pdf"), width = 12, height = 12)
Heatmap(cluster7_mat, name = "Expression",  
        column_split = factor(cluster7_cluster_anno),
        cluster_columns = TRUE,
        show_column_dend = FALSE,
        cluster_column_slices = TRUE,
        column_title_gp = gpar(fontsize = 12),
        column_gap = unit(0.5, "mm"),
        cluster_rows = TRUE,
        show_row_dend = FALSE,
        col = col_fun,
        row_names_gp = gpar(fontsize = 12),
        column_title_rot = 90,
        top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = scales::hue_pal()(22)),
                                           labels_gp = gpar(col = "white", fontsize = 10))),
        show_column_names = FALSE,
        use_raster = TRUE,
        raster_quality = 4)
dev.off()

# KLRB1 = CD161
FeaturePlot(cd8_minus_annotated_cluster7, features = c("TRDV2", "TRGV9", "KLRD1", "KLRB1"), order = TRUE)
ggsave(here("results/figures/cd8_minus_cluster7_featureplot.pdf"), width = 8, height = 4)
```


### cell type percentage by groups

Relate to Main Figure 2b,d,f,g. Supp Figure 6

```{r}
cd8_minus_annotated<- readRDS(here("data/cd8_minus_annotated_subcluster7_seurat.rds"))
## total number of cells in the whole cd8- subset
#number_cells_per_sample<- cd8_minus_annotated@meta.data %>%
#  group_by(tigl_id) %>%
#  summarise(total_cells = n()) %>%
#  ungroup()

## use the total number of cells per sample from the whole cd3+ subset as denominator

cd3_plus_annotated<- readRDS(here("data/cd3_plus_annotated_seurat_subcluster.rds"))
number_cells_per_sample<- cd3_plus_annotated@meta.data %>%
  group_by(tigl_id) %>%
  summarise(total_cells = n()) %>%
  ungroup()


## total 26 clusters
clusters<- cd8_minus_annotated@meta.data %>% 
  pull(clusters_anno_sub) %>%
  unique() %>%
  sort()

## total 64 samples
samples<- cd8_minus_annotated@meta.data %>%
  select(tigl_id, group) %>%
  distinct() 

## total should be 64 * 26
cluster_levels<- tidyr::expand_grid(samples, clusters_anno = clusters) %>%
  mutate(cluster_levels = paste(tigl_id, group, clusters_anno, sep = "|")) %>%
  pull(cluster_levels)

number_cells_per_seurat_cluster<- cd8_minus_annotated@meta.data %>%
  mutate(clusters_anno = paste(tigl_id, group, clusters_anno_sub, sep = "|")) %>%
  mutate(clusters_anno= factor(clusters_anno, levels = cluster_levels )) %>%
  count(clusters_anno,name = "cells_each_cluster",  .drop = FALSE) %>% 
  tidyr::separate(col = clusters_anno, into =c("tigl_id", "group", "clusters_anno"), sep = "\\|") 

seurat_cells_percentage<- number_cells_per_seurat_cluster %>%
  left_join(number_cells_per_sample) %>%
  mutate(percentage = cells_each_cluster/total_cells) %>%
  mutate(group = factor(group, levels = c("Healthy_donors", "New_C1D1", "refractory_CR_C1D1","refractory_PR_C1D1", "refractory_PD_C1D1","refractory_CR_C4D1","refractory_PR_C4D1", "refractory_PD_C4D1" )))

seurat_cells_percentage$clusters_anno

write_csv(seurat_cells_percentage, here("results/cd8_minus_seurat_annotated_cluster_percentage_subcluster7_cd3plus_total_denominator.csv"))


ggplot(seurat_cells_percentage, aes(x = group, y = percentage)) +
  geom_boxplot(aes(fill = group), outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1) +
  scale_y_continuous(labels = scales::percent) + 
  facet_wrap(~clusters_anno, scales = "free") +
  theme_bw(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_text(size = 8)) +
  xlab("")

ggsave(here("results/figures/cd8_minus_seurat_annotaed_cluster_percentage_boxplot_fix_0_subcluster7_cells_cd3plus_total_denominator.pdf"), width = 16, height = 10)

ggplot(seurat_cells_percentage, aes(x = group, y = percentage)) +
  geom_beeswarm(aes(color = group), size = 0.5) +
  scale_y_continuous(labels = scales::percent) + 
  facet_wrap(~clusters_anno, scales = "free") +
  theme_bw(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_text(size = 8)) +
  xlab("")


```

cuzick trend test https://www.rdocumentation.org/packages/PMCMRplus/versions/1.4.4/topics/cuzickTest
to se if there is a increasing trend in cell percentage in a cluster for pre-CR--> pre-PR -->pre-PD

```{r}
library(tidyr)
#install.packages("PMCMRplus")
library(PMCMRplus)
seurat_cells_percentage<- read_csv(here("results/cd8_minus_seurat_annotated_cluster_percentage_subcluster7_cd3plus_total_denominator.csv"))

## pre-treatment
cells_percentage_nest<- seurat_cells_percentage %>% 
  filter(group %in% c("refractory_CR_C1D1", "refractory_PR_C1D1", "refractory_PD_C1D1")) %>%
  mutate(group = factor(group, levels =c("refractory_CR_C1D1", "refractory_PR_C1D1", "refractory_PD_C1D1"))) %>%
  group_by(clusters_anno) %>%
  nest()

## post-treatment
cells_percentage_nest<- seurat_cells_percentage %>% 
  filter(group %in% c("refractory_CR_C4D1", "refractory_PR_C4D1", "refractory_PD_C4D1")) %>%
  mutate(group = factor(group, levels =c("refractory_CR_C4D1", "refractory_PR_C4D1", "refractory_PD_C4D1"))) %>%
  group_by(clusters_anno) %>%
  nest()

cuzickTest(cells_percentage_nest$data[[1]]$percentage, cells_percentage_nest$data[[1]]$group)

cells_percentage_nest<- cells_percentage_nest %>%
  mutate(cuzick_test = map_dbl(data, ~ cuzickTest(.x$percentage, .x$group)$p.value))

cells_percentage_nest$cuzick_test

View(cells_percentage_nest %>% select(-data)) 
### healthy donor vs newly diagnosed

cells_percentage_nest<- seurat_cells_percentage %>% 
  filter(group %in% c("Healthy_donors", "New_C1D1")) %>%
  mutate(group = factor(group, levels =c("Healthy_donors", "New_C1D1"))) %>%
  group_by(clusters_anno) %>%
  nest()

wilcox.test(percentage~ group, data = cells_percentage_nest$data[[1]]) 
cells_percentage_nest<- cells_percentage_nest %>%
  mutate(wilcox_test = map_dbl(data, function(x) wilcox.test(percentage ~group, data=x)$p.value))

  
seurat_cells_percentage %>% 
  group_by(group, clusters_anno) %>%
  summarise(median_percentage = median(percentage)) %>%
  write_tsv(here("data/cd8_minus_seurat_annotated_cluster_median_percentage_subcluster7_cd3plus_total_denominator.csv"))


seurat_cells_percentage %>% 
  distinct(tigl_id, group, total_cells) %>%
  arrange(group, total_cells) %>%
  mutate(tigl_id = factor(tigl_id, levels = .$tigl_id) ) %>%
  ggplot(aes(x= tigl_id, y = total_cells)) +
  geom_bar(stat ="identity",aes(fill = group)) +theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("")

ggsave(here("results/figures/cd8_minus_final_seurat_obj_cells_number.pdf"), width =12, height = 6)


```

